R Python, and Ruby clients for GBIF species occurrence data

### 2017-09-06 manuscript.tex
added Ruby client and fixes use case problems, updated circleci badge in readme

\documentclass[author-year, review, 11pt]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
\makeatother

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
\newcommand{\RegionMarkerTok}[1]{{#1}}
\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
\newcommand{\NormalTok}[1]{{#1}}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={R and Python clients for GBIF species occurrence data},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{R and Python clients for GBIF species occurrence data}
    \author[cstar]{Scott Chamberlain\corref{c1}}
   \ead{scott(at)ropensci.org} 
   \cortext[c1]{Corresponding author}
    \author[boettig]{Carl Boettiger}
   \ead{carl(at)ropensci.org}

@@

  Email address:
  \href{mailto:scott@ropensci.org}{\href{mailto:scott@ropensci.org}{\nolinkurl{scott@ropensci.org}}}

@@

  Methods. We have developed a client to access GBIF data for each of the
  R and Python programming languages: \texttt{rgbif} and \texttt{pygbif}.

  Results. For both clients, we describe their design and utility, and
  demonstrate many use cases.

@@

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. Biodiversity
records have been used for a suite of other use cases: validating
habitat suitability models with real occurrence data (Ficetola et al.,
2014); ancestral range reconstruction (Ferretti et al., 2015; Mar{í}a
Mendoza et al., 2015); development of invasive species watch lists
(Faulkner et al., 2014); evaluating risk of invasive species spread
(Febbraro et al., 2013); and effects of climate change on future
biodiversity (Brown et al., 2015).

@@

There are many online services that collect and maintain specimen
records. However, Global Biodiversity Information Facility (hereafter,
GBIF, \href{http://www.gbif.org}{\url{http://www.gbif.org}}) is the
largest collection of biodiversity records globally, currently with 643
million records, 1.6 million taxa, 15,450 datasets from 780 publishers
(as of 2016-02-09). Many large biodiversity warehouses such as
iNaturalist
(\href{http://www.inaturalist.org}{\url{http://www.inaturalist.org}}),
VertNet (\href{http://vertnet.org}{\url{http://vertnet.org}}), and
USGS's Biodiversity Information Serving Our Nation (BISON;
\href{http://bison.usgs.ornl.gov}{\url{http://bison.usgs.ornl.gov}}) all
feed into GBIF.

The most important data organizational level in GBIF is the individual
record. In R using \texttt{rgbif} you'll recognize this as a row in the
data.frame that is returned from \texttt{occ\_search()} or
\texttt{occ\_data()}. Going upstream, each record is part of a dataset,
where each dataset is submitted by an organization, organizations are
organized into nodes, datasets are published through institutions (which
may be hosted at another organization), and a network is a group of
datasets (managed by GBIF).

Each record has some taxonomic name associated with it, which itself is
linked to a lot of other taxonomic data. GBIF maintains their own
taxonomic data.

This organization matters because you can navigate through GBIF data
through these various organizational levels.

@@

@@

The \texttt{rgbif} package is nearly completely written in R (a small
Javascript library is included for reading well known text (Herring,
2011)), uses an \href{http://choosealicense.com/licenses/mit/}{MIT
license} to maximize use everywhere. \texttt{rgbif} is developed
publicly on GitHub at
\href{https://github.com/ropensci/rgbif}{\url{https://github.com/ropensci/rgbif}},
where development versions of the package can be installed, and bugs and
feature requests reported. Stable versions of \texttt{rgbif} can be
installed from
\href{https://cran.rstudio.com/web/packages/rgbif/}{CRAN}, the
distribution network for R packages. \texttt{rgbif} is part of the
rOpenSci project (\url{http://ropensci.org}), a developer network making
R software to facilitate reproducible research.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{import} \NormalTok{pygbif}
\end{Highlighting}
\end{Shaded}

@@

@@

\texttt{rgbif} and \texttt{pygbif} are designed following the
\href{http://www.gbif.org/developer/summary}{GBIF Application
Programming Interface}, or API. The GBIF API has four major components:
registry, taxonomic names, occurrences, and maps. We also include
functions to interface with the OAI-PMH GBIF service; only dataset
information is available vis this service, however. We ignore maps in
both libraries as it is concerned with generating maps primarily for web
applications. Both libraries have a suite of functions dealing with each
of registry, taxonomic names, and occurrences - we'll go through each in
turn describing design and example usage.

@@

With each request \texttt{rgbif} and \texttt{pygbif} make to GBIF's API,
we send request headers that tell GBIF what client the request is coming
from, including what version of the library. This helps GBIF know what
proportion of requests are coming from which client, and therefore from
R vs.~Python; this information is helpful for GBIF in thinking about how
people are using GBIF data.

@@

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Datasets
\item
  Installations
\item
  Networks
\item
  Nodes
\item
  Organizations
\end{itemize}

Dataset information in general is available via the OAI-PMH service,
functions in \texttt{rgbif} prefixed with \texttt{gbif\_oai\_}, but not
available in \texttt{pygbif} yet.

@@

Dataset functions include search, dataset metadata retrieval, and
dataset metrics. Searching for datasets is an important part of the
discovery process. One can search for datasets on the GBIF web portal.
However, programmatic searching using this package is much more
powerful. Identifying datasets appropriate for a research question is
helpful as you can get metadata for each dataset, and track down dataset
specific problems, if any.

The \texttt{dataset\_search()} function is one way to search for
datasets. Here, we search for the query term ``oregon'', which finds any
datasets that have terms matching that term.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{dataset_search}\NormalTok{(}\DataTypeTok{query =} \StringTok{"oregon"}\NormalTok{)}
\NormalTok{res$data$datasetTitle[}\DecValTok{1}\NormalTok{:}\DecValTok{10}\NormalTok{]}
\CommentTok{#>  [1] "UCDavis - Western USA - Monarch Butterflies - 1892-2005"                                                                         }
\CommentTok{#>  [2] "A geographic distribution database of Mononychellus mites (Acari: Tetranychidae) on cassava (Manihot esculenta)"                 }
\CommentTok{#>  [3] "Plantas Acuáticas de la Orinoquía Colombiana"                                                                                    }
\CommentTok{#>  [4] "USBombus, contemporary survey data of North American bumble bees (Hymenoptera,  Apidae, Bombus) distributed in the United States"}
\CommentTok{#>  [5] "Wool carder bees of the genus Anthidium in the Western Hemisphere"                                                               }
\CommentTok{#>  [6] "CM Birds Collection"                                                                                                             }
\CommentTok{#>  [7] "SDNHM Birds Collection"                                                                                                          }
\CommentTok{#>  [8] "University of British Columbia Herbarium (UBC) - Bryophytes Collection"                                                          }
\CommentTok{#>  [9] "University of British Columbia Herbarium (UBC) - Vascular Plant Collection"                                                      }
\CommentTok{#> [10] "Bryophyte Collection - University of Washington Herbarium (WTU)"}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{pygbif }\CharTok{import} \NormalTok{registry}
\NormalTok{registry.datasets(}\DataTypeTok{type}\NormalTok{=}\StringTok{"OCCURRENCE"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Dataset metrics are another useful way of figuring out what datasets you
may want to use. One drawback is that these metrics data are only
available for datasets of type \emph{checklist}, but there are quite a
lot of them (16084).

Here, we search for dataset metrics for a single dataset, with uuid
\texttt{ec93a739-1681-4b04-b62f-3a687127a17f}, a checklist of the ants
(Hymenoptera: Formicidae) of the World.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{dataset_metrics}\NormalTok{(}\DataTypeTok{uuid=}\StringTok{'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{rank =} \KeywordTok{names}\NormalTok{(res$countByRank),}
           \DataTypeTok{count =} \KeywordTok{unname}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(res$countByRank)))}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[c]{@{}lr@{}}
\toprule
rank & count\tabularnewline
\midrule
\endhead
SPECIES & 13710\tabularnewline
SUBSPECIES & 3234\tabularnewline
GENUS & 726\tabularnewline
TRIBE & 53\tabularnewline
SUBFAMILY & 20\tabularnewline
FAMILY & 2\tabularnewline
KINGDOM & 1\tabularnewline
PHYLUM & 1\tabularnewline
CLASS & 1\tabularnewline
ORDER & 1\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{pygbif }\CharTok{import} \NormalTok{registry}
\NormalTok{registry.dataset_metrics(uuid=}\StringTok{'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{networks}\NormalTok{(}\DataTypeTok{limit =} \DecValTok{10}\NormalTok{)$data$title}
\CommentTok{#>  [1] "GBIF Backbone Sources"                                      }
\CommentTok{#>  [2] "Canadensys"                                                 }
\CommentTok{#>  [3] "Southwest Collections of Arthropods Network (SCAN)"         }
\CommentTok{#>  [4] "VertNet"                                                    }
\CommentTok{#>  [5] "Dryad"                                                      }
\CommentTok{#>  [6] "GBIF Network"                                               }
\CommentTok{#>  [7] "The Knowledge Network for Biocomplexity (KNB) "             }
\CommentTok{#>  [8] "Online Zoological Collections of Australian Museums (OZCAM)"}
\CommentTok{#>  [9] "Catalogue of Life"                                          }
\CommentTok{#> [10] "Ocean Biogeographic Information System (OBIS)"}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CharTok{from} \NormalTok{pygbif }\CharTok{import} \NormalTok{registry}
\NormalTok{registry.networks(limit = }\DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

and in Python

>>>>

\documentclass[author-year, review, 11pt]{components/elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}
\usepackage{lineno} % add
  \linenumbers % turns line numbering on
\bibliographystyle{elsarticle-harv}
\biboptions{sort&compress} % For natbib
\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\it \hfill\today}%
 \let\@evenfoot\@oddfoot}
 \def\tightlist{}
\makeatother

@@

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={R Python, and Ruby clients for GBIF species occurrence data},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{R Python, and Ruby clients for GBIF species occurrence data}
    \author[cstar]{Scott Chamberlain\corref{c1}}
   \ead{scott(at)ropensci.org} 
   \cortext[c1]{Corresponding author}
    \author[boettig]{Carl Boettiger}
   \ead{carl(at)ropensci.org}

@@

  Email address:
  \href{mailto:scott@ropensci.org}{\nolinkurl{scott@ropensci.org}}

@@

  Methods. We have developed clients to access GBIF data for each of the
  R, Python, and Ruby programming languages: \texttt{rgbif},
  \texttt{pygbif}, \texttt{gbifrb}.

  Results. For all clients we describe their design and utility, and
  demonstrate many use cases.

@@

Data on individuals, including which species, and where they're found,
can be used for a large number of research questions. Biodiversity
records have been used for a suite of other use cases: validating
habitat suitability models with real occurrence data (Ficetola et al.,
2014); ancestral range reconstruction (Ferretti et al., 2015; María
Mendoza et al., 2015); development of invasive species watch lists
(Faulkner et al., 2014); evaluating risk of invasive species spread
(Febbraro et al., 2013); and effects of climate change on future
biodiversity (Brown et al., 2015).

@@

There are many online services that collect and maintain specimen
records. However, Global Biodiversity Information Facility (hereafter,
GBIF, \url{http://www.gbif.org}) is the largest collection of
biodiversity records globally, currently with 820 million records,
roughly 5.9 million taxa, 36,000 datasets from 1,300 publishers (as of
2016-02-09). Many large biodiversity warehouses such as iNaturalist
(\url{http://www.inaturalist.org}), VertNet (\url{http://vertnet.org}),
and USGS's Biodiversity Information Serving Our Nation (BISON;
\url{http://bison.usgs.ornl.gov}) all feed into GBIF.

The most important organizational level in GBIF occurrence data is the
occurrence record. The fields in a record vary, but include information
about taxonomy (kingdom, phylum, genus, species names) and their
identifiers, dataset metadata, and locality information including
geospatial position. Going upstream, each record is part of a dataset,
where each dataset is submitted by an organization, organizations are
organized into nodes, datasets are published through institutions (which
may be hosted at another organization), and a network is a group of
datasets (managed by GBIF).

Each occurrence record has some taxonomic name associated with it, which
itself is linked to a lot of other taxonomic data - including a master
taxonomic backbone that integrates taxonomies across many taxonomic
authorities.

The organization of GBIF matters because you can navigate GBIF data
through these hierarchical organizational levels - it helps to be
familiar with the terminology and how each group relates to another.

@@

Although we discuss libraries for R, Python, and Ruby here, we focus
mostly on the R library \texttt{rgbif} as it has seen the most developer
and user attention, and is the most mature.

@@

The \texttt{rgbif} package is nearly completely written in R (a small
Javascript library is included for reading well known text (Herring,
2011)), uses an \href{http://choosealicense.com/licenses/mit/}{MIT
license} to maximize use everywhere. \texttt{rgbif} is developed
publicly on GitHub at \url{https://github.com/ropensci/rgbif}, where
development versions of the package can be installed, and bugs and
feature requests reported. Stable versions of \texttt{rgbif} can be
installed from
\href{https://cran.rstudio.com/web/packages/rgbif/}{CRAN}, the
distribution network for R packages. \texttt{rgbif} is part of the
rOpenSci project (\url{http://ropensci.org}), a developer network making
R software to facilitate reproducible research.

@@

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{import}\NormalTok{ pygbif}
\end{Highlighting}
\end{Shaded}

@@

\subsection{gbifrb}\label{gbifrb}

\texttt{gbifrb} (Chamberlain) is a library for working with GBIF data in
the Ruby programming environment. Like Python, Ruby is a general purpose
programming language used widely in all sectors. Unlike Python, Ruby is
not used extensively in scientific disciplines. However, a Ruby client
for GBIF data can be an important tool given how widely Ruby is used for
web and web service development.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{gem install gbifrb}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{require }\StringTok{'gbifrb'}
\end{Highlighting}
\end{Shaded}

The \texttt{gbifrb} library is less mature and complete than the R and
Python libraries. It also uses an
\href{http://choosealicense.com/licenses/mit/}{MIT license} to maximize
use everywhere. \texttt{gbifrb} is developed publicly on GitHub at
\url{https://github.com/sckott/gbifrb}, where development versions of
the package can be installed, and bugs and feature requests reported.
Stable versions of \texttt{gbifrb} can be installed from
{[}Rubygems{]}{[}gemgbif{]}, the distribution network for Ruby
libraries.

@@

\texttt{rgbif}, \texttt{pygbif}, and \texttt{gbifrb} are designed
following the \href{http://www.gbif.org/developer/summary}{GBIF
Application Programming Interface}, or API. The GBIF API has four major
components: registry, taxonomic names, occurrences, and maps. We also
include functions to interface with the OAI-PMH GBIF service; only
dataset (registry) information is available via this service, however.
An interface to the GBIF maps API is in development for \texttt{rgbif},
but is non-existent for both \texttt{pygbif} and \texttt{gbifrb}. All
three libraries have a suite of functions dealing with each of registry,
taxonomic, names, and occurrences - we'll go through each in turn
describing design of the user interface and example usage.

@@

With each request \texttt{rgbif}, \texttt{pygbif}, \texttt{gbifrb} make
to GBIF's API, we send request headers that tell GBIF what library the
request is coming from, including what version of the library. This
helps GBIF know what proportion of requests are coming from which
library, and therefore from R vs.~Python vs.~Ruby; this information is
helpful for GBIF in thinking about how people are using GBIF data.

@@

\begin{itemize}
\tightlist
\item
  Datasets
\item
  Installations
\item
  Networks
\item
  Nodes
\item
  Organizations
\end{itemize}

Dataset information in general is available via the OAI-PMH service,
functions in \texttt{rgbif} prefixed with \texttt{gbif\_oai\_}, but not
available in \texttt{pygbif} or \texttt{gbifrb} yet.

@@

Dataset functions include search, dataset metadata retrieval, and
dataset metrics. Searching for datasets is an important part of the
discovery process. One can search for datasets on the GBIF web portal.
However, programmatic searching using any of these libraries is more
powerful. Identifying datasets appropriate for a research question is
helpful as you can get metadata for each dataset, and track down dataset
specific problems, if any.

The \texttt{dataset\_search()} function in \texttt{rgbif} is one way to
search for datasets. Here, we search for the term ``oregon'', which
finds any datasets that have words matching that term.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{dataset_search}\NormalTok{(}\DataTypeTok{query =} \StringTok{"oregon"}\NormalTok{)}
\NormalTok{res}\OperatorTok{$}\NormalTok{data}\OperatorTok{$}\NormalTok{datasetTitle[}\DecValTok{1}\OperatorTok{:}\DecValTok{10}\NormalTok{]}
\CommentTok{#>  [1] "Oregon State Ichthyology Collection"                                                                                                                                                  }
\CommentTok{#>  [2] "Oregon State University Herpetological Collection"                                                                                                                                    }
\CommentTok{#>  [3] "Mygalomorph spiders from southwestern Oregon, USA, with descriptions of four new species"                                                                                             }
\CommentTok{#>  [4] "A new species of Helobdella (Hirudinida: Glossiphoniidae) from Oregon, USA"                                                                                                           }
\CommentTok{#>  [5] "Annotated Checklist of the large branchiopod crustaceans of Idaho, Oregon and Washington, USA, with the “ rediscovery ” of a new species of Branchinecta (Anostraca: Branchinectidae)"}
\CommentTok{#>  [6] "A new species of Chrysobothris Eschscholtz from Oregon and Washington, with notes on other Buprestidae (Coleoptera) occurring in the United States and Canada"                        }
\CommentTok{#>  [7] "Three new species of Grylloblatta Walker (Insecta: Grylloblattodea: Grylloblattidae), from southern Oregon and northern California"                                                   }
\CommentTok{#>  [8] "A new species of Cladotanytarsus (Lenziella) from Oregon supports the systematic concept of the subgenus (Diptera: Chironomidae)"                                                     }
\CommentTok{#>  [9] "A new monster from southwest Oregon forests: Cryptomasterbehemoth sp. n. (Opiliones, Laniatores, Travunioidea)"                                                                       }
\CommentTok{#> [10] "Two new species of Fluminicola (Caenogastropoda, Lithoglyphidae) from southwest Oregon, USA, and a range extension for F. multifarius"}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ pygbif }\ImportTok{import}\NormalTok{ registry}
\NormalTok{registry.datasets(}\BuiltInTok{type}\OperatorTok{=}\StringTok{"OCCURRENCE"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

In Ruby, we can do the same. Here, search for datasets of type
\texttt{OCCURRENCE}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{require }\StringTok{'gbifrb'}
\NormalTok{registry = }\DataTypeTok{Gbif}\NormalTok{::}\DataTypeTok{Registry}
\NormalTok{registry.datasets(}\StringTok{type: "OCCURRENCE"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

Dataset metrics are another useful way of figuring out what datasets you
may want to use. One drawback is that these metrics data are only
available for datasets of type \emph{checklist}, but there are quite a
lot of them (21502).

Here, in R we search for dataset metrics for a single dataset, with uuid
\texttt{ec93a739-1681-4b04-b62f-3a687127a17f}, a checklist of the ants
(Hymenoptera: Formicidae) of the World.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res <-}\StringTok{ }\KeywordTok{dataset_metrics}\NormalTok{(}\DataTypeTok{uuid=}\StringTok{'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{rank =} \KeywordTok{names}\NormalTok{(res}\OperatorTok{$}\NormalTok{countByRank),}
           \DataTypeTok{count =} \KeywordTok{unname}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(res}\OperatorTok{$}\NormalTok{countByRank)))}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}lr@{}}
\toprule
rank & count\tabularnewline
\midrule
\endhead
SPECIES & 13710\tabularnewline
SUBSPECIES & 3234\tabularnewline
GENUS & 726\tabularnewline
TRIBE & 53\tabularnewline
SUBFAMILY & 20\tabularnewline
FAMILY & 2\tabularnewline
KINGDOM & 1\tabularnewline
PHYLUM & 1\tabularnewline
CLASS & 1\tabularnewline
ORDER & 1\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ pygbif }\ImportTok{import}\NormalTok{ registry}
\NormalTok{registry.dataset_metrics(uuid}\OperatorTok{=}\StringTok{'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

The same in Ruby:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{require }\StringTok{'gbifrb'}
\NormalTok{registry = }\DataTypeTok{Gbif}\NormalTok{::}\DataTypeTok{Registry}
\NormalTok{registry.dataset_metrics(}\StringTok{uuid: 'ec93a739-1681-4b04-b62f-3a687127a17f'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{networks}\NormalTok{(}\DataTypeTok{limit =} \DecValTok{10}\NormalTok{)}\OperatorTok{$}\NormalTok{data}\OperatorTok{$}\NormalTok{title}
\CommentTok{#>  [1] "GBIF Backbone Sources"                                      }
\CommentTok{#>  [2] "Canadensys"                                                 }
\CommentTok{#>  [3] "Southwest Collections of Arthropods Network (SCAN)"         }
\CommentTok{#>  [4] "VertNet"                                                    }
\CommentTok{#>  [5] "Dryad"                                                      }
\CommentTok{#>  [6] "GBIF Network"                                               }
\CommentTok{#>  [7] "The Knowledge Network for Biocomplexity (KNB) "             }
\CommentTok{#>  [8] "Online Zoological Collections of Australian Museums (OZCAM)"}
\CommentTok{#>  [9] "Catalogue of Life"                                          }
\CommentTok{#> [10] "Ocean Biogeographic Information System (OBIS)"}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\ImportTok{from}\NormalTok{ pygbif }\ImportTok{import}\NormalTok{ registry}
\NormalTok{registry.networks(limit }\OperatorTok{=} \DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

And in Ruby:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{require }\StringTok{'gbifrb'}
\NormalTok{registry = }\DataTypeTok{Gbif}\NormalTok{::}\DataTypeTok{Registry}
\NormalTok{registry.networks(}\StringTok{limit: }\DecValTok{10}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

And in Python
